A Formalization of Typed Lua

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: new version (RC5)

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; V & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

@@

The type {\bf void} is a type with no values, used by the
type system as a way to detect branches that are unreachable
due to flow typing. This type is neutral with regards to
the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.

@@

The subtyping rules are a deduction system for the
subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
The variable $\senv$ is a set of subtyping assumptions used
by the subtyping relation between recursive types. 
Most of the subtyping rules are standard, so we will
focus on the interesting cases.

@@

The type environment is standard, and maps
identifiers to first-level types. Some typing rules of
Section~\ref{sec:tables} can change the type of an identifier,
so the typing relation has both an input and an output type
environment. The {\em projection environment} $\Pi$ maps projection labels to second-level types, and we explain
its use in Section~\ref{sec:projections}.

@@

\begin{figure*}
	{\small
	\[
	\begin{array}{rcl}
	wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{F:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
    vt(L, V) & = & fix(V) \\
    vt(F_1, F_2) & = & nil(fix(F_2))\\
    vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
    nil(F) & = & F \;\;\; \mathrm{if}\;\Nil \lesssim F\\
    nil(F) & = & F \cup \Nil \;\;\; \mathrm{otherwise}\\\\
	fix(\{\overline{F:V}\}_{unique|open}) & = &
	\{\overline{F:V}\}_{fixed}\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^\prime}\}_{unique},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}
	\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_c V_r \,\vee\, V^r \lesssim_c V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique|open},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
	\{\overline{F:sup_c(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{open}
	\\
	merget(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	joint(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:nil(V^\prime)},\overline{F^{\prime\prime}:nil(V^{\prime\prime})}\}_{unique}\\
	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \lesssim_u V_2\\
	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \lesssim_u V_1\\\\
	reopen(\{\overline{F:V}\}_{closed}) & = & \{\overline{F:V}\}_{open} \\
	reopen(F) & = & F
	\end{array}
	\]}
	\caption{Auxiliary Functions}
\label{fig:funcs}
\end{figure*}

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
This is not enough, as assignment
to a closed-over variable can be unsafe. In the following
example, assume that the $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}$:
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}

@@

	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\

@@

proj(F \times P, i) & = & proj(P, i-1)

@@

A standard way of decomposing the union in the assignment
would let the type of $q$be $\Integer \times \Nil$ and
the type of $r$ be $\Integer \times \String$. Flow typing
would narrow the type of $q$ to $\Integer$ in the {\bf then}
branch and to $\Nil$ in the {\bf else} branch, but would
leave the type of $r$ unchanged.

@@

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. The more interesting cases are inside the
branches of the {\bf if} statement, where rule {\sc T-IfFilterNilProj} filters the whole union of tuples referenced
by the projection type.

@@

Assignment to an l-value with a projection type is forbidden. Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
the projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}

>>>>

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; F & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
%K ::= & & \textsc{{\small key types:}}\\
%& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

@@

@@

The subtyping rules are a deduction system for the
subtyping relation $\subtype$ between two types of the
same kind.
The variable $\senv$ is a set of subtyping assumptions used
by the subtyping relation between recursive types. 
Most of the subtyping rules are standard, so we will
focus on the interesting cases.

@@

The type environment is standard, and maps
identifiers to expression types. Some typing rules of
Section~\ref{sec:tables} can change the type of an identifier,
so the typing relation has both an input and an output type
environment. The {\em projection environment} $\Pi$ maps projection labels to second-level types, and we explain
its use in Section~\ref{sec:projections}.

@@

\begin{figure*}
	{\small
	\[
	\begin{array}{rcl}
	wf(\{\overline{F:V}\}_{unique|open|fixed|closed}) & = & \forall i \; ((\nexists j \; i \not= j \,\wedge\, F_{i} \lesssim F_{j}) \,\wedge\, wf(V_{i}) \,\wedge\\
	& & \;\;\;\; \lnot tag(V_{i},unique) \,\wedge\, \lnot tag(V_{i},open))\\
	wf({\bf const}\; F) & = & wf(F) \\
	wf(F_1 \cup F_2) & = & wf(F_1) \,\wedge\, wf(F_2) \\
	wf(\mu x.F) & = & wf(F) \\
	wf(S_1 \rightarrow S_2) & = & wf(S_1) \,\wedge\, wf(S_2) \\
	wf(S_1 \sqcup S_2) & = & wf(S_1) \,\wedge\, wf(S_2)\\
	wf(F{*}) &=& wf(F) \\
	wf(F \times P) &=& wf(F) \,\wedge\,wf(P)\\
	wf(F) & = & \top \;\;\;\mathrm{for\; all\; other\; cases}\\\\
	tag(F_1 \cup F_2, t) & = & tag(F_1, t) \,\vee\, tag(F_2,t) \\
	tag(\{\overline{F:V}\}_t, t) & = & \top\\
	\multicolumn{3}{c}{
		\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{other\;recusive\;cases\;are\;similar,\;other\;base\;cases\;are\;}\bot}\\
\\\\
    vt(L, V) & = & fix(V) \\
    vt(F_1, F_2) & = & nil(fix(F_2))\\
    vt(F_1, {\bf const}\;F_2) & = & {\bf const}\;nil(fix(F_2))\\\\
    nil(T) & = & T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
    nil(T) & = & T \cup \Nil \;\;\; \mathrm{otherwise}\\\\
	fix(\{\overline{F:V}\}_{unique|open}) & = &
	\{\overline{F:V}\}_{fixed}\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\mathrm{other\;cases\;by\;straightforward\;structural\;recursion}}\\\\
	merge(\overline{\env}) & = & reduce(\overline{\env}, merge2)\\
	merge2(\env_1,\env_2) & = & \{\overline{(id,merget(\env_1(id),\env_2(id))}\}\\
	merget(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
	merget(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^\prime}\}_{unique},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}
	\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_c V_r \,\vee\, V^r \lesssim_c V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	merget(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique|open},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique|open}) &=&
	\{\overline{F:sup_c(V^l,V^r)},\overline{F^\prime:V^\prime},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{open}
	\\
	merget(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	join(\env_1,\env_2) & = & \{\overline{(id,joint(\env_1(id),\env_2(id))}\}\\
	joint(T_1, T_2) & = & T_1 \;\;\; \mathrm{if\;} T_2 \lesssim T_1\\
	joint(T_1, T_2) & = & T_2 \;\;\; \mathrm{if\;} T_1 \lesssim T_2\\
	\multicolumn{3}{c}{\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\mathrm{the\;next\;case\;applies\;if}\;
		\overline{V^l \lesssim_u V_r \,\vee\, V^r \lesssim_u V_l}\;\mathrm{and\;the\;right\; side\; is\;}wf}\\
	joint(\{\overline{F:V^l},\overline{F^{\prime}:V^{\prime}}\}_{unique},
	\{\overline{F:V^r},\overline{F^{\prime\prime}:V^{\prime\prime}}\}_{unique}) &=&
	\{\overline{F:sup_u(V^l,V^r)},\overline{F^\prime:nil(V^\prime)},\overline{F^{\prime\prime}:nil(V^{\prime\prime})}\}_{unique}\\
	joint(T_1, T_2) & = & \bot \;\;\; \mathrm{otherwise}\\\\
	sup_u(V_1, V_2) & = & V_2 \;\;\; \mathrm{if}\; V_1 \lesssim_u V_2\\
	sup_u(V_1, V_2) & = & V_1 \;\;\; \mathrm{if}\; V_2 \lesssim_u V_1\\\\
	reopen(\{\overline{F:V}\}_{closed}) & = & \{\overline{F:V}\}_{open} \\
	reopen(F) & = & F
	\end{array}
	\]}
	\caption{Auxiliary Functions}
\label{fig:funcs}
\end{figure*}

@@

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
This is not enough, as assignment
to a closed-over variable can be unsafe. In the following
example, assume that $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}_{closed}$:
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; \lfloor g({<}\{``x":
			 \String \cup \Nil\}_{closed}{>} \; a) \rfloor_{0}$}
	\end{tabular}
\end{center}

@@

	\env_{1}, \penv \vdash el:E, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto proj(E, 1), \ldots, id_{n} \mapsto proj(E, n)]\\

@@

proj(F \times P, i) & = & proj(P, i-1)\\
proj(E{*}, i) & = & nil(E) \\
proj(T \times E, 1) & = & T \\
proj(T \times E, i) & = & proj(E, i-1)

@@

A standard way of decomposing the union in the assignment
would let the type of $q$ be $\Integer \times \Nil$ and
the type of $r$ be $\Integer \times \String$. Flow typing
would narrow the type of $q$ to $\Integer$ in the {\bf then}
branch and to $\Nil$ in the {\bf else} branch, but would
leave the type of $r$ unchanged.

@@

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$ via rule {\sc T-IdReadProj}, which just uses auxiliary function $proj$. The more interesting cases are inside the
branches of the {\bf if} statement, where rule {\sc T-IfFilterNilProj} filters the whole union of tuples referenced
by the projection type.

@@

Assignment to an l-value with a projection type is forbidden
(there is no rule {\sc T-IdWriteProj}). Allowing assignment is unsound, as it can break the dependency among the components of the union. In the example below,
the projected type of $r$ outside of the {\bf if} statement
is $\Integer \cup \String$, so the assignment seems correct,
but inside $s_1$ the projected type of $r$ would still be
$\Integer$:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(a, b) \, \mathbf{in}$} \\
		&  $r = ``foo" ;\; \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$
	\end{tabular}
\end{center}
